---
title: "[Tsaroucha, A., Rathcke. T. & Canzi. M.] Effects of a face mask on the perception of English fricatives by native speakers of Greek"
date: "2021-03-31"
tags: ["acoustics", "phonetics", "perception"]
weight: 2
draft: false
output: 
  html_document:
    fig.align = "center"
---

### Presentations

__...__

### Abstract

```{r setup, include = F}
library(car) 
library(emmeans) 
library(ggeffects)
library(lme4) 
library(lmerTest) 
library(tidyverse)
```

```{r readr, include = F}
read_csv("L2-Perception.csv") %>%
  filter(L1 == "Greek") %>%
  rename(ENG.level = ENG.current.lever) %>%
  mutate(RT = log(RT), Syllable = ifelse(Syllable == "onset", "Onset", "Coda"),
         COG = scale(COG), Intensity = scale(Intensity), 
         ENG.lextale = scale(ENG.lextale), 
         Competitor = as.factor(Competitor),
         Accuracy = ifelse(Accuracy == TRUE, "1", "0"),
         Device = ifelse(Device == "computer", "Computer", "Mobile"),
         Visual = as.factor(Visual), Participant.ID = as.factor(Participant.ID)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate_if(is.logical, as.factor) %>% 
  within(ENG.level <- relevel(x = ENG.level, ref = "intermediate")) -> Greek
```

__Background.__ The ongoing _COVID-19_ pandemic has made a daily use of face masks mandatory. This has greatly influenced the everyday communication since face masks interfere with both acoustic and visual cues that would otherwise be available to interlocutors in spoken interaction. The present study focuses on the effect of this interference for the perception of 4 voiceless English fricatives: dental /θ/, labiodental /f/, alveolar /s/, and postalveolar /ʃ/. It investigates Greek listeners’ responses to these fricatives produced with and without a face mask to examine whether L2 perception suffers in face mask conditions. The Greek phonology does not include a fricative contrast between an alveolar and a postalveolar place of articulation, with the Greek /s/ being classified as a retracted alveolar (Arvaniti, 2007), i.e. located in-between the English alveolar and postalveolar fricatives (Panagopoulos, 1991). Previous research on the perception of English consonants by native listeners of Greek has demonstrated that the alveolar-postalveolar pair causes a great difficulty to these listeners since Greek lacks the contrast, while in the dental-labiodental pair, /f/ tends to be accurately identified.

__Hypotheses.__ Following on from the work by Lengeris and Nicolaidis (2014; 2016), the present study hypothesized that the identification of /s/ and /ʃ/ would be greatly impaired in face mask conditions, while the dental-labiodental pair was expected to cause less difficulties to native Greek listeners. We further hypothesized an impaired fricative identification in coda positions (esp. for dentals and labiodentals) given that Greek phonotactics only allows for /s/ and /n/ to occur in syllable codas (Setatos, 1974). English proficiency was expected to increase the overall accuracy, esp. in the identification of /s/ and /ʃ/.

__Method.__ Forty-five healthy Greek listeners aged 18-50 (36F) participated in an audio-visual phoneme monitoring task conducted online (Connine & Titone, 1996). Participants’ level of L2-English was B1 or higher (CERF, 2001); their lexical L2-proficiency was 52.50% or above (Lemhöfer & Broersma, 2012).

__Results.__ A series of mixed-effects models was fit to the data. Best-fit models of accuracy revealed an effect of the face mask condition, but for the dental-labiodental pair only, with higher error rates in the mask-on condition. A significant three-way interaction (Figure 1) indicated that the identification of masked /f/ and /θ/ was impaired particularly in the coda position. The presence of a face mask affected reaction times in response to the dental-labiodental pair but not to alveolar-postalveolar pair (Figure 2). Higher L2 proficiency improved the identification of /s/ only, regardless of the mask condition.

__Figure 1.__ _Model plot of the three-way interaction of condition * fricative * syllable structure displaying accuracy estimates for the perception of the target fricatives in no-mask (blue) vs. mask (red) condition in onset (left) vs. coda (right) position._

```{r, include = F}
ggpredict(Greek %>% mutate(Target = fct_relevel(Target, "F", "TH", "S", "SH"),
                           Syllable = fct_relevel(Syllable, "Onset", "Coda")) %>%
            mutate_if(is.character, as.factor) %>%
            glmer(formula = Accuracy ~ Condition * Target * Syllable +
                    LogFreq : Condition + ENG.lextale * Target +
                    (1 + Visual.nm + Condition | Participant.ID) + (1 | Spelling), 
                  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
                  family = "binomial"), 
          terms = c("Target", "Condition", "Syllable"), type = "fe") -> Acc.predict
```

```{r, echo = F}
Acc.predict %>% ggplot(aes(x, predicted, 
                        colour = group, pch = group, linetype = group)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_errorbar(ymin = Acc.predict$conf.low, ymax = Acc.predict$conf.high, width = 0.2, size = 0.4) +
  scale_colour_manual(values = c("deepskyblue4", "indianred3"), 
                      name = "Condition", labels = c("No Mask", "Mask")) +
  scale_shape_manual(values = c(16, 15), name = "Condition", labels = c("No Mask", "Mask")) +
  scale_linetype_manual(values = c(1, 2), name = "Condition", labels = c("No Mask", "Mask")) +
  theme_minimal() +
  ylim(0.5, 1) +
  xlab("Target Fricative") + ylab("Estimates of Accuracy") +
  facet_wrap(~facet)
```

__Figure 2.__ _Reaction time estimates for the fricative perception in no-mask (blue) vs. mask (red) condition grouped by the presence of visual cues._

```{r, include = F}
ggpredict(Greek %>% filter(Accuracy == "1") %>%
            mutate(Target = fct_relevel(Target, "F", "TH", "S", "SH")) %>%
  lmer(formula = RT ~ Condition * Visual.nm + Syllable + Device +
         LogFreq + (1 + Condition | Participant.ID) + (1 | Spelling) + (1 | Browser)), 
  terms = c("Visual.nm", "Condition"), type = "fe") %>% 
  mutate(predicted = exp(1) ^ predicted, conf.low = exp(1) ^ conf.low,
         conf.high = exp(1) ^ conf.high) -> RT.predict
```

```{r, echo = F}
RT.predict %>% ggplot(aes(x, predicted, 
                        colour = group, pch = group, linetype = group)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_errorbar(ymin = RT.predict$conf.low, ymax = RT.predict$conf.high, width = 0.2, size = 0.4) +
  scale_colour_manual(values = c("deepskyblue4", "indianred3"), 
                      name = "Condition", labels = c("No Mask", "Mask")) +
  scale_shape_manual(values = c(16, 15), name = "Condition", labels = c("No Mask", "Mask")) +
  scale_linetype_manual(values = c(1, 2), name = "Condition", labels = c("No Mask", "Mask")) +
  theme_minimal() +
  ylim(700, 1500) +
  xlab("Presence of Visual Cues") + ylab("Estimates of Reaction Times (ms)") +
  theme(aspect.ratio = 1/1)
```

__Conclusions.__ Overall, these results corroborate previous findings by Llamas et al. (2008), suggesting that speech with face coverings is less intelligible due to the lack of visual information. This finding further speaks to the idea that misperceptions of the dental-labiodental pair might result from the acoustic and articulatory similarities of the two English sounds rather than from L1-interference (Lengeris & Nicolaidis, 2014). Overall, the study demonstrated that face coverings during speech poses challenges to L2-perception primarily due to the lack of visual cues which cannot be offset by a high L2-proficiency.

__References.__

* Arvaniti, A. (2007). Greek Phonetics The State of the Art. Journal of Greek linguistics, 8(1), 97-208. 
* Connine, C. M. & Titone, D. (1996). Phoneme Monitoring. Language and Cognitive Processes, 11(6), 635-645.
* Lemhöfer, K., & Broersma, M. (2012). Introducing LexTALE: A quick and valid lexical test for advanced learners of English. Behavior research methods, 44(2), 325-343.
* Lengeris, A., & Nicolaidis, K. (2014). English consonant confusions by Greek listeners in quiet and noise and the role of phonological short-term memory. In Fifteenth Annual Conference of the International Speech Communication Association.
* Lengeris, A., & Nicolaidis, K. (2016). The identification and production of English consonants by Greek speakers. Selected papers on theoretical and applied linguistics, 21, 211-223.
* Llamas, C., Harrison, P., Donnelly, D. & Watt, D. (2008). Effects of different types of face coverings on speech acoustics and intelligibility. York Papers in
Linguistics (Series 2) 9, 80–104.
* Setatos, M. (1974). Phonology of Modern Greek [in Greek]. Athens: Papazisis Editions. 
* Panagopoulos, L. (1991). A comparison of English and Greek alveolar fricatives. In Proceedings of the XIIth International Congress of Phonetic Sciences, Vol. 2, pp. 326-328.

```{r, include = F}
sessionInfo()
```